A look into Kamala Harris' polls in the upcoming 2024 US Presidential Election 

### 2024-11-04 paper/paper.qmd
Edited data, added all graphs and tables

library(tidyverse)
library(kableExtra)
library(rstanarm)

just_harris_high_quality <-
  read_csv(file = here::here("data/02-analysis_data/harris.csv"))

bayesian_model <-
  readRDS(file = here::here("models/electionmodel.rds"))

spline_model <- 
  readRDS(file = here::here("models/splinemodel.rds"))
```

@@

The dataset provided by fivethirtyeight contains over 15000 observations across 50 variables. Each observation is a poll conducted on the 2024 US Presidential Election. Pollsters are required by fivethirtyeight to disclose certain information about their poll, either publicly, or through an investigation. Of this information, a handful are key to the predictive model.

@@

### **`poll_score:`**

The error and bias associated with a pollster. Positive values indicate more bias and errors, negative values indicate less. This value will tell us if biases in polls influences the predicted percentages of each candidate

```{r}
#| include: true
#| warning: false
#| message: false

#### Plot data ####
base_plot <- ggplot(just_harris_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date")

# Color by pollscore
base_plot +
  geom_point(aes(color = factor(pollscore))) +
  geom_smooth() +
  theme(legend.position = "bottom")

pollscore_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(pollscore, na.rm = TRUE),
    Mean = mean(pollscore, na.rm = TRUE),
    Median = median(pollscore, na.rm = TRUE),
    Max = max(pollscore, na.rm = TRUE)
  )

pollscore_summary %>% 
  kable() %>%
  kable_styling()

@@

### `sample_size`

Polls are conducted over a subset of the targeted population (Hispanics, young people, Texans, LGBTQ+, rural communities, etc.).

```{r}
#| include: true
#| warning: false
#| message: false

#### Plot data ####
base_plot +
  geom_point(aes(color = factor(sample_size))) +
  geom_smooth() +
  theme(legend.position = "bottom")

@@

### `state`

The state that each poll as conducted in. If there are any empty values, we assume these are national polls.

```{r}
#| include: true
#| warning: false
#| message: false

#### Plot data ####
just_harris_high_quality %>%
  ggplot(mapping=aes(x = state), y=count(n)) +
  geom_bar()

state_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(count(state), na.rm = TRUE),
    Max = max(state, na.rm = TRUE)
  )

state_summary %>% 
  kable() %>%
  kable_styling()

@@

```{r}
#| include: true
#| warning: false
#| message: false

#### Plot data ####
base_plot <- ggplot(just_harris_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date")

# Plots poll estimates and overall smoothing
base_plot +
  geom_point() +
  geom_smooth()

# Color by pollster
# This gets messy - need to add a filter - see line 21
base_plot +
  geom_point(aes(color = pollster)) +
  geom_smooth() +
  theme(legend.position = "bottom")

# Facet by pollster
# Make the line 21 issue obvious
# Also - is there duplication???? Need to go back and check
base_plot +
  geom_point() +
  geom_smooth() +
  facet_wrap(vars(pollster))

# Color by pollscore
base_plot +
  geom_point(aes(color = factor(pollscore))) +
  geom_smooth() +
  theme(legend.position = "bottom")

@@

### `party_binary`

@@

```{r}
#| include: true
#| warning: false
#| message: false

#### Plot data ####
just_harris_high_quality %>%
  ggplot(mapping=aes(x = end_date)) +
  geom_histogram()

end_date_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(end_date, na.rm = TRUE),
    Max = max(end_date, na.rm = TRUE)
  )

end_date_summary %>% 
  kable() %>%
  kable_styling()

>>>>

library(tidyverse)
library(kableExtra)
library(rstanarm)
library(dplyr)

# Read in data from clean_data.r
just_harris_high_quality <-
  read_csv(file = here::here("data/02-analysis_data/analysis_data_harris.csv"))

just_trump_high_quality <-
  read_csv(file = here::here("data/02-analysis_data/analysis_data_trump.csv"))
```

@@

The dataset provided by fivethirtyeight contains over 15000 observations across 50 variables. Each observation is a poll conducted on the 2024 US Presidential Election. Fivethirtyeight's dataset was chosen due to it's comprehensive review of polls and pollsters, attributing grades in error and bias, as well as in transparency. Their thorough investigations ensure their dataset not only contains insightful polls, but also includes as much relevant information about the polls conducted as possible. This gives us plenty of predictors when building a model that will caluclate the percentage win of either Trump or Harris.

The dataset was read into R [@citeR], and cleaned using the R package tidyverse [@tidyverse]. Due to Kamal's late entry into the race, some polls do not have any data on her specifically. This is why we set a cut-off date at 2024/07/21, the day Kamala replaced Joe Biden as the Democratic nominee.

Fivethirtyeight includes a metric called `numeric_grade`, which combines the transparency_score and transparency score. We only included polls with a `numeric_grade` at or above 2.7. This gave us the best balance between quantity and quality. Finally, we create 2 datasets, one with polls predicting a Harris win, the other a Trump win.

#### Create end_date - pct base plots for Trump and Harris ####
base_plot_kamala <- ggplot(just_harris_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date")

base_plot_trump <- ggplot(just_trump_high_quality, aes(x = end_date, y = pct)) +
  theme_classic() +
  labs(y = "Trump percent", x = "Date")
```

@@

### **`transparency_score`**

Fivethirtyeight awards a score out of 10 to each pollster based on how informative their methodology is. Understanding the methodology of a pollster assits in understanding any errors or biases present in their findings.

```{r}
#| label: fig-transscore_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their transparency_score, which is fivethirtyeight's metric for transpareny in the pollster's methodology. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(transparency_score))) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-transscore_harris
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Kamala Harris win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Harris
transparency_score_harris_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Displaying summary
transparency_score_harris_summary %>% 
  kable() %>%
  kable_styling()
```

```{r}
#| label: fig-transscore_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trump's win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their transparency_score, which is fivethirtyeight's metric for transpareny in the pollster's methodology.
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_trump +
  geom_point(aes(color = factor(transparency_score))) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-transscore_trump
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Donald Trump win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Trump
numeric_grade_trump_summary <- just_trump_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Display summary
transparency_score_trump_summary %>% 
  kable() %>%
  kable_styling()

@@

### **`numeric_grade`**

A cumulative score given to pollster that combines fivethirtyeight's `pollscore` and `transparency_score`. Like `transparency_score`, we believe this metric can give us insight into potential biases in each pollster and poll, which is why it is included in our model.

```{r}
#| label: fig-numgrade_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their numeric_grade, which is fivethirtyeight's combined metric on transparency and bias/errors. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(numeric_grade))) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```
```{r}
#| label: tbl-numgrade_harris
#| tbl-cap: Summary of numeric_grades amongst polls predicting a Kamala Harris win.
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Harris
numeric_grade_harris_summary <- just_harris_high_quality %>%
  summarize(
    Min = min(numeric_grade, na.rm = TRUE),
    Mean = mean(numeric_grade, na.rm = TRUE),
    Median = median(numeric_grade, na.rm = TRUE),
    Max = max(numeric_grade, na.rm = TRUE)
  )

# Display summary
numeric_grade_harris_summary %>% 
  kable() %>%
  kable_styling()

@@

```{r}
#| label: fig-numgrade_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trumps' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on their numeric_grade, which is fivethirtyeight's combined metric on transparency and bias/errors. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = factor(numeric_grade))) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```
```{r}
#| label: tbl-transscore_trump
#| tbl-cap: Summary of transparency_scores amongst polls predicting a Donald Trump win. The closer the values are to 10, the more trustworthy the predictions. 
#| echo: false
#| warning: false
#| message: false

# Generating summary statistics for Trump
numeric_grade_trump_summary <- just_trump_high_quality %>%
  summarize(
    Min = min(transparency_score, na.rm = TRUE),
    Mean = mean(transparency_score, na.rm = TRUE),
    Median = median(transparency_score, na.rm = TRUE),
    Max = max(transparency_score, na.rm = TRUE)
  )

# Display summary
numeric_grade_trump_summary %>% 
  kable() %>%
  kable_styling()

@@

### `pollster`

```{r}
#| label: fig-pollster_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Harris' win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on the pollster who authored the poll. 
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_kamala +
  geom_point(aes(color = pollster)) +
  geom_smooth(color = "blue") +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-pollster_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. The polls in this graph predict Trump's win percentage, with the average in the blue line. The dots in the graph below represent individual polls conducted in the months leading up to the election. They are coloured based on the pollster who authored the poll.  
#| echo: false
#| warning: false
#| message: false

# Color by transparency_score
base_plot_trump +
  geom_point(aes(color = pollster)) +
  geom_smooth(color = "red") +
  theme(legend.position = "bottom")
```

### `states`

```{r}
#| label: fig-state
#| fig-cap: Battleground states are what usually determines the winner of a presidential election. This figure outlines the polling outcomes of each candidate from August to November, for each of the seven battleground states this election cycle.
#| echo: false
#| warning: false
#| message: false
#| 
# Define major swing states
swing_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")

combined_data <- bind_rows(
  just_harris_high_quality |> mutate(candidate = "Kamala Harris"),
  just_trump_high_quality |> mutate(candidate = "Donald Trump")
)

# Filter data for Harris and Trump in the swing states
swing_states_data <- combined_data |>
  filter(state %in% swing_states)

# Create a base plot for swing states data
swing_states_plot <- ggplot(swing_states_data, aes(x = end_date, y = pct, color = candidate)) +
  theme_minimal() +
  labs(x = "Polling End Date", y = "Polling Percentage", title = "Polling Outcomes in Swing States") +
  theme(legend.position = "bottom")

# Facet the plot by state
swing_states_plot +
  geom_point(aes(shape = candidate), alpha = 0.7) +
  geom_smooth(aes(linetype = candidate), se = FALSE) +
  facet_wrap(vars(state)) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))

@@

@@

```{r}
#| label: fig-enddate_harris
#| fig-cap: Polls are conducted from a start date to an end date. The closer the end date is to election day, the more insightful the results are to analysts and the general public. This distribution shows the end date of polls predicting a Kamala Harris win.
#| echo: false
#| warning: false
#| message: false

#### Plot data ####
just_harris_high_quality %>%
  ggplot(mapping=aes(x = end_date)) +
  geom_histogram(fill = "blue", alpha = 0.7, color = "black",)
```

```{r}
#| label: fig-enddate_trump
#| fig-cap: Polls are conducted from a start date to an end date. The closer the end date is to election day, the more insightful the results are to analysts and the general public. This distribution shows the end date of polls predicting a Donald Trump win.
#| echo: false
#| warning: false
#| message: false

#### Plot data ####
just_trump_high_quality %>%
  ggplot(mapping=aes(x = end_date)) +
  geom_histogram(fill = "red", alpha = 0.7, color = "black",)
```
